---
title: "mtcars-regression"
author: "Nick"
date: "Tuesday, September 22, 2015"
output: html_document
---

Load data

```{r}
data(mtcars)
```

Create linnear model with mpg as the coutcome for all variables

```{r}
fit <- lm(mpg ~ . , data = mtcars)
summary(fit)$coefficients
```

Create linnear model with mpg as the outcome with cyl as predictor

```{r}
fit_mpg_cyl <- lm(mpg ~ cyl , data = mtcars)
summary(fit_mpg_cyl)$coefficients
```

Create linnear model with mpg as the outcome with cyl as predictor & weight as confounder

```{r}
fit_mpg_cyl_wt <- lm(mpg ~ cyl + wt, data = mtcars)
summary(fit_mpg_cyl_wt)$coefficients
```

Examine the means by cyl

```{r}
library(dplyr)
summarise(group_by(mtcars, cyl), mn = mean(mpg))
```

Treat the cyl as a factor with 4 a s base

```{r}
cyl2 <- relevel(as.factor(mtcars$cyl), "4")
summary(lm(mpg ~ cyl2 + wt, data = mtcars))$coef

```

Compare difference between 4 & 8 cyl with and without weight variable

```{r}
summary(lm(mpg ~ cyl2 + wt, data = mtcars))$coef[3,1]
summary(lm(mpg ~ cyl2, data = mtcars))$coef[3,1]
```

Compare the above model with a second that considers and interaction between cyl and weight

```{r}
fitA <- lm(mpg ~ cyl2 + wt, data = mtcars)
fitB <- lm(mpg ~ cyl2 + wt + wt * cyl2, data = mtcars)
anova(fitA, fitB)
```

Scale the weight variable by 0.5
```{r}
lm(mpg ~ I(wt * 0.5) + factor(cyl), data = mtcars)
```

Look at most influential points
```{r}
x <- c(0.586, 0.166, -0.042, -0.614, 11.72)
y <- c(0.549, -0.026, -0.127, -0.751, 1.344)
fit_x_y <- lm(y ~ x)
hatvalues(fit_x_y)
```

Find the dfbeta for the point with the highest hat value

```{r}
dfbetas(fit_x_y)[5,2]
```