---
title: "mtcars-assignement"
author: "Nick"
date: "Wednesday, September 23, 2015"
output: html_document
---

### Context

You work for Motor Trend, a magazine about the automobile industry. Looking at a data set of a collection of cars, they are interested in exploring the relationship between a set of variables and miles per gallon (MPG) (outcome). They are particularly interested in the following two questions:

"Is an automatic or manual transmission better for MPG"

"Quantify the MPG difference between automatic and manual transmissions"

## Executive Summary

## Detailed Analysis

First lets just have a cursory look at the first few entries in our dataset:
```{r echo=FALSE}
data(mtcars)
head(mtcars)
```

Here we see some examples of the data and what we can expect to work with.  There are a total of `r ncol(mtcars)` variables and we also know there are a total of `r nrow(mtcars)` observations.

```{r echo=FALSE}
mpgVsTrans <- data.frame(trans=as.factor(mtcars$am), mpg=mtcars$mpg)
levels(mpgVsTrans$trans) <- c('Automatic', 'Manual')

automatic.mean <- round(mean(mpgVsTrans[mpgVsTrans$trans == 'Automatic', 2]),2)
manual.mean <- round(mean(mpgVsTrans[mpgVsTrans$trans == 'Manual', 2]),2)
means.dat <- data.frame(trans=levels(mpgVsTrans$trans), means=c(automatic.mean, manual.mean))

library(ggplot2)
g <- ggplot(mpgVsTrans, aes(x = mpg, fill = trans))
g <- g + geom_histogram(colour = "black", binwidth=2)
g <- g + geom_vline(aes(xintercept=means), data=means.dat, size=2)
g <- g + facet_grid(. ~ trans)
g <- g + coord_fixed(ratio=5)
```

An inital examination of the data shows that automatic and manual transmissions have a mean of `r automatic.mean` and `r manual.mean` mpg respectively. We can also see the data is fairly sparse given the gaps in the distribution.

```{r fig.width=7, fig.height=3, fig.align='center', echo=FALSE}
g
```

```{r}
fit <- lm(mpg ~ am , data = mtcars);
plot(mpgVsTrans, main="Transmission vs MPG");
```
   

## Limitations of Study

There data itself has not mechanism to cross reference our data, many factors can play a role in the consumption of fuel.  Different tire brands compounts, the fuel octane, driver or rolling road used.  This information is not captured.

Ultimately comparing only mpg vs transmission type is likely a trivialisation of the data resulting in *underfitting* we're omitting necessary covariates resulting in *bias*.  Unfortunately with the data at hand fitting more covariates results in large variance thus a less useful model. 

The ideal way to conduct the study would be to have makes and models of cars that have both a manual and automatic variant to pair like with like. Then with all other variables being even we could answer with a higher degree of certainty.
