---
title: "Coursera - Statistical Inference Project"
author: "Nicholas Smith"
date: "Friday, July 24, 2015"
output: html_document
---

## Overview

This project will investigate the exponential distribution in R and compare it with the Central Limit Theorem. The exponential distribution can be simulated in R with rexp(n, $\lambda$) where $\lambda$ is the rate parameter.


### 1) Show the sample mean and compare it to the theoretical mean of the distribution.

Consider the mean of the Exponential Distribution is 1/$\lambda$

For this analysis we will be using a lambda value of 0.2

```{r}
lambda <- 0.2
lambda
```

Therefore our expected mean will be 5

```{r}
expected_mean <- 1/lambda 
expected_mean
```

First we need to simulate some data, we will run a 1000 simulations determining the average of 40 exponentials.

```{r}
sample_means = NULL
for (i in 1 : 1000){
  sample_means = c(sample_means, mean(rexp(40, lambda)))  
} 
```

The centre of our distrubtion is therefore:

```{r}
sample_mean <- mean(sample_means)
sample_mean
```

```{r}
hist(sample_means)
```

### 2. Show how variable the sample is (via variance) and compare it to the theoretical variance of the distribution.

### 3) Show that the distribution is approximately normal

